Performance benefits of Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the parallel processing of large neuroimaging data

### 2020-03-21 paper.tex
some text fixes

\newcommand{\valerie}[1]{\color{blue}\textbf{Note from Valerie}:
      #1 \color{black}}
\newcommand{\tristan}[1]{\color{orange}\textbf{Note from Tristan}:
      #1 \color{black}}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached petabyte scale, and continue to grow.
The ability to leverage the entirety of these datasets is limited to a restricted
number of labs with both the capacity and infrastructure to
process the data. Whereas Big Data engines have significantly
reduced application performance penalties with respect to data
movement, their applied strategies (e.g. data locality, in-memory computing and lazy evaluation)
are not necessarily practical
within neuroimaging workflows where intermediary results may
need to be materialized to shared storage for post-processing
analysis. In this paper we evaluate the performance advantage
brought by Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the
processing of large neuroimaging datasets using the two available
configurations modes: Memory mode and App Direct mode. We
employ a synthetic algorithm on the 76 GiB and 603 GiB BigBrain, as well as apply a
standard neuroimaging application on the Consortium for Reliability
and Reproducibility (CoRR) dataset using 25 and 96 parallel
processes in both cases. Our results show that the performance of
applications leveraging persistent memory is superior to that of other
storage devices, with the exception of DRAM. This is the case
in both Memory and App Direct mode and irrespective of the
amount of data and parallelism. Furthermore, persistent memory in App
Direct mode is believed to benefit from the use of DRAM as
a cache for writing when output data is significantly
smaller than available memory. We believe the use of persistent memory
will be beneficial to both neuroimaging applications running on
HPC or visualization of large, high-resolution images.
\end{abstract}

@@

Furthermore, intermediary data is often required for post-processing analysis. This limits
any performance benefits that can arise from volatile in-memory computing as intermediary
results need to be materialized onto persistent storage. As neuroimaging datasets continue
to increase in size, the movement of data will significantly increase the processing time.

To mitigate the effects of data writes, the Linux kernel has implemented
strategies, such as the writeback cache. The writeback cache
allows processes to use the memory as a cache for writing. This strategy contrasts writethrough, which 
writes data directly to the device. The writeback cache size
is configurable, but nevertheless limited. When writes approach the cache's capacity,
processes performing writes start to be throttled. Once the cache capacity is reached,
processes can no longer use the cache until all cached written data is flushed
to the appropriate storage device.

Whereas Operating Systems and popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. writeback cache, in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing persistent storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

@@

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed. In order to quantify the effects of storage devices
on such a dataset, we have implemented a basic synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cached
in-memory, by ensuring that read and written data are not the same. This application
was parallelized in two different ways: using both GNU Parallel~\cite{gnuparallel} and Apache Spark 2.4.3 (PySpark)~\cite{spark}. 
All code was implemented in Python 3.6.

There are some notable differences between the GNU Parallel and Apache Spark implementations.
For instance, in the GNU Parallel implementation, all operations (i.e. reading, incrementing and writing)
are done within a single task. This task is applied to all the \bigbrain blocks using GNU Parallel,
processing a subset of them at a time,
depending on the level of parallelism provided. Reading in the GNU Parallel implementation 
is achieved using the popular neuroimaging I/O library NiBabel v2.5.0~\cite{nibabel}. When provided with a filename, as is the case
for the GNU Parallel implementation, Nibabel will simply load the header in memory, and memory map the data
using \href{https://numpy.org/}{NumPy}. This step will be referred to as ``load header''. It is only when the data is actually required
(i.e. during incrementation), that the data will be loaded into memory.

@@

It is believed that Optane DCPMM can be useful for a variety of neuroimaging applications. For
instance, typically applications executing in HPC environments can benefit from 
the speedups provided by Optane DCPMM. Furthermore, image visualization servers will be able
to display higher resolution images at significantly greater speeds.
\section{Acknowledgement}
We are very grateful to the Intel and Dell-EMC teams for giving us the
opportunity to experiment with the Intel Optane DC Server, for providing
high-quality technical support, and for their thorough feedback on the
manuscript. 
\bibliographystyle{IEEEtran} 
\bibliography{biblio}

>>>>

\newcommand{\valerie}[1]{\color{blue}\textbf{Note from Valerie}:
      #1 \color{black}}
\newcommand{\tristan}[1]{\color{orange}\textbf{Note from Tristan}:
      #1 \color{black}}
\captionsetup{belowskip=-1pt}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached petabyte scale and continue to grow.
The ability to leverage the entirety of these datasets is limited to a restricted
number of research laboratories with both the capacity and infrastructure to
process the data. Whereas Big Data engines have significantly
reduced application performance penalties with respect to data
movement, their applied strategies (e.g. data locality, in-memory computing and lazy evaluation)
are not necessarily practical
within neuroimaging workflows where intermediary results may
need to be materialized to shared storage for post-processing
analysis. In this paper we evaluate the performance advantage
brought by Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the
processing of large neuroimaging datasets using the two available
configurations modes: Memory mode and App Direct mode. We
employ a synthetic algorithm on the 76 GiB and 603 GiB BigBrain, as well as apply a
standard neuroimaging application on the Consortium for Reliability
and Reproducibility (CoRR) dataset using 25 and 96 parallel
processes in both cases. Our results show that the performance of
applications leveraging persistent memory is superior to that of other
storage devices, with the exception of DRAM. This is the case
in both Memory and App Direct mode, irrespective of the
amount of data and parallelism. Furthermore, persistent memory in App
Direct mode is believed to benefit from the use of DRAM as
a cache for writing when output data is significantly
smaller than available memory. We believe the use of persistent memory
will be beneficial to both neuroimaging applications running on
HPC or visualization of large, high-resolution images.
\end{abstract}

@@

Intermediary data is often required for post-processing analysis. This limits
any performance benefits that can arise from volatile in-memory computing as intermediary
results need to be materialized onto persistent storage. As neuroimaging datasets continue
to increase in size, the movement of data will significantly increase the processing time.

To mitigate the effects of data writes, the Linux kernel has implemented
strategies, such as the writeback cache. The writeback cache
allows processes to use the memory as a cache for writing. This strategy contrasts writethrough, which 
writes data directly to the device. The writeback cache size
is configurable, but nevertheless limited. When writes approach the cache's capacity,
processes performing writes begin to be throttled. Once the cache capacity is reached,
processes can no longer use the cache until all cached written data is flushed
to the appropriate storage device.

Whereas operating systems (OS) and popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. writeback cache, in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing persistent storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

@@

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed. In order to quantify the effects of storage devices
on such a dataset, we have implemented a basic synthetic application that splits an image into blocks and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cached
in-memory, by ensuring that read and written data are not the same. This application
was parallelized in two different ways: using both GNU Parallel~\cite{gnuparallel} and Apache Spark 2.4.3 (PySpark)~\cite{spark}. 
All code was implemented in Python 3.6.

There are some notable differences between the GNU Parallel and Apache Spark implementations.
For instance, in the GNU Parallel implementation, all operations (i.e. reading, incrementing and writing)
are done within a single task. This task is applied to all the \bigbrain blocks using GNU Parallel,
which then processes a subset of them at a time
depending on the level of parallelism provided. Reading in the GNU Parallel implementation 
is achieved using the popular neuroimaging I/O library NiBabel v2.5.0~\cite{nibabel}. When provided with a filename, as is the case
for the GNU Parallel implementation, Nibabel will simply load the header in memory, and memory map the data
using \href{https://numpy.org/}{NumPy}. This step will be referred to as ``load header''. It is only when the data is actually required
(i.e. during incrementation), that the data will be loaded into memory.

@@

It is believed that Optane DCPMM can be useful for a variety of neuroimaging applications. For
instance, typically applications executing in HPC environments can benefit from 
the speedups provided by Optane DCPMM. Furthermore, image visualization servers will be able
to display higher resolution images at significantly greater speeds.
\section{Acknowledgement}
We are very grateful to the Intel and Dell-EMC teams for giving us the
opportunity to experiment with the Intel Optane DC Server, for providing
high-quality technical support, and, in  particular, Steve Scargall, for
providing thorough technical support on the manuscript.
\bibliographystyle{IEEEtran} 
\bibliography{biblio}
